A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-02 tex/codatadp.tex
more

@@

\paragraph{Conventions.}
This paper assumes a monomorphic (or many-sorted) first-order logic throughout.
A signature $\Sig = (\Types, \Funcs)$ consists of a set of types $\Types$ and a
set of function symbols $\Funcs$. Types are simply atomic sorts, with no
structure, and interpreted by nonempty domains. The set $\Types$ must contain a
distinguished type \ty{bool} interpreted as the set of truth
values. %, and may contain other interpreted types (e.g., \ty{int}, \ty{real}).
Predicates are simply functions that return $\ty{bool}$.
The metavariables $\delta, \tau$ range over types, with $\delta$ reserved
for (co)datatypes.

Function symbols are written in a sans-serif font (e.g., $\const{f}$, $\const{g}$) to
distinguish them from variables (e.g., $x$, $y$). With each function symbol \const{f}
is associated a list of argument types $\tau_1,\ldots,\tau_n$ (with $n \ge 0$)
and a return type $\tau$. This connection can be expressed compactly as
$\const{f} : \tau_1 \times \cdots \times \tau_n \to \tau$
(which collapses to $\const{f} : \tau$ if $n = 0$).
For a term $t$, the notation $t : \tau$ indicates that $t$ has type $\tau$.
Functions are applied the standard way, with $\const{f}(t_1,\ldots,t_n)$
applying the $n$-ary function symbol
\const{f} to $n$ well-typed arguments $t_1 : \tau_1$, \ldots, $t_n :
\tau_n$. Nullary function symbols are called constants and can appear without
parentheses in terms.

@@

Axiomatization:

  * axiomatically
    * injectivity
      distinctness
      exhaustiveness
    * datatypes:
      * induction axiom (schema)
        * consequence: no infinite objects (in particular, no cyclic objects)

@@

Datatypes can be defined semantically as the initial model of an equational
specification of the selector--constructor equations. This is the view followed
by Barrett et al.\ \cite{barrett-et-al-2010}. A disadvantage with this approach
is that it does not naturally account for selectors applied to wrong
constructors. Barrett et al.\ address this by parameterizing the construction by
default values, but these give rise to spurious equalities between unrelated
terms---for example, $\const{s}_{11}(\const{C}_2) \teq \const{s}_{11}(\const{C}_3)$.
This could be corrected, but the added
compexity seems to suggest that selectors are better specified axiomatically.

>>>>

\newcommand\jPrime{j\vthinspace'\negvthinspace}

@@

\paragraph{Conventions.}
This paper assumes a monomorphic (or many-sorted) first-order logic throughout.
A signature $\Sig = (\Types, \Funcs)$ consists of a set of types $\Types$ and a
set of function symbols $\Funcs$. Types are simply atomic sorts, with no
structure, and interpreted by nonempty domains. The set $\Types$ must contain a
distinguished type \ty{bool} interpreted as the set of truth
values. %, and may contain other interpreted types (e.g., \ty{int}, \ty{real}).
The only predicate is $\teq$, and it is treated as a logical symbol,
just like the connectives and quantifiers of first-order logic.
Other predicates can be represented as functions that return $\ty{bool}$.
The metavariables $\delta, \tau$ range over types, with $\delta$ reserved
for datatypes and codatatypes.

Function symbols are written in a sans-serif font (e.g., $\const{f}$, $\const{g}$) to
distinguish them from variables (e.g., $x$, $y$). Upper case names (e.g., $\const{C}$,
$\const{Nil}$) are reserved for constructors. With each function symbol \const{f}
is associated a list of argument types $\tau_1,\ldots,\tau_n$ (with $n \ge 0$)
and a return type $\tau$. This connection can be expressed compactly as
$\const{f} : \tau_1 \times \cdots \times \tau_n \to \tau$
(which collapses to $\const{f} : \tau$ if $n = 0$).
For a term $t$, the notation $t : \tau$ indicates that $t$ has type $\tau$.
Functions are applied the standard way, with $\const{f}(t_1,\ldots,t_n)$
applying the $n$-ary function symbol
\const{f} to $n$ well-typed arguments $t_1 : \tau_1$, \ldots, $t_n :
\tau_n$. Nullary function symbols are called constants and can appear without
parentheses in terms.
%
Finally, the notation $\bar x$ abbreviates a list or tuple $x_1,\ldots,x_n$.

@@

Datatypes and codatatypes can be characterized axiomatically. They share basic
properties of constructors and selectors. All properties below range over all
$i$, $j$, $\jPrime$, and $k$ within bounds and to all possible splits of the
$n$-ary constructor $\const{C}_{i\negvthinspace j}$'s argument list into $\bar x,
y, \bar z$:
%
\[
\begin{aligned}[t]
\text{Distinctness:}\quad
  & \forall \bar x, \bar y.\;\, \const{C}_{i\negvthinspace j}(\bar x) \tneq \const{C}_{i\negvthinspace \jPrime}(\bar y) \qquad\text{if $j \not= \jPrime$}
  \\
\text{Injectivity:}\quad
  & \forall \bar x, y, \bar z.\;\, \const{C}_{i\negvthinspace j}(\bar x, y, \bar z) \teq \const{C}_{i\negvthinspace j}(\bar x, y', \bar z) \longrightarrow y \teq y'
  \\
\text{Exhaustiveness:}\quad
  &
  \\
\text{Selection:}\quad
  &
\end{aligned}
\]

    * datatypes:
      * induction axiom (schema)
        * consequence: no infinite objects (in particular, no cyclic objects)

@@

Datatypes can be defined semantically as the initial model of an equational
specification of the selector--constructor equations. This is the view followed
by Barrett et al.\ \cite{barrett-et-al-2010}. A disadvantage with this approach
is that it does not naturally account for selectors applied to wrong
constructors. Barrett et al.\ address this by parameterizing the construction by
default values, but these give rise to spurious equalities between unrelated
terms---for example, $\const{s}_{11}(\const{C}_2) \teq \const{s}_{11}(\const{C}_3)$.
This could be corrected, but the added
complexity seems to suggest that selectors are better specified axiomatically.
